---
title: "project"
format: pdf
---
```{r}
library(dplyr)
library(tidyverse)
```

```{r}
DonorHyperglycemia <- read_csv("DonorHyperglycemia.csv")
```


***REAL*** testing donor glucose association with dysfunction & differing by risk index
```{r}

# Full model (with interaction: glucose × donor risk)
m1 <- glm(delayed_fn ~ log2(glutwa) * donorrisk + d_genderf,
          data = DonorHyperglycemia,
          family = "binomial")

summary(m1)

# Reduced model (no interaction)
m0 <- glm(delayed_fn ~ log2(glutwa) + donorrisk + d_genderf,
          data = DonorHyperglycemia,
          family = "binomial")

summary(m0)

# Likelihood Ratio Test (comparing both models)
anova(m0, m1, test = "LRT")

```


 Donor Glucose: 3.63x higher odds of graft dysfunction (earlier - proven to not be statistically significant)
 Donor Risk: 410x higher odds of graft dysfunction for each 1 point increase (proven to be statistically significant), but at the interecept (when glucose = 1, which is not realistic. So we will run another )
 Gender: 0.95 (5% lower odds) of graft dysfunction for females (not statistically signficant)
 


```{r}
library(ggplot2)

ggplot(DonorHyperglycemia, aes(x = donorrisk, y = delayed_fn)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(x = "Donor Risk Index", y = "Probability of Graft Dysfunction")
```

donor risk association with dysfunction (and adjusted)
```{r}
m <- glm(delayed_fn ~ donorrisk + d_genderf,
         data = DonorHyperglycemia,
         family = "binomial")

summary(m)
```
donorrisk → estimate = 0.977, p = 0.00281 (**statistically significant)

odds ratio 
```{r}
exp(coef(m))
```

CI for the odds ratio
```{r}
exp(confint(m))
```
We are 95% confident that a one-unit increase in donor risk is associated with between 1.4× and 5.1× higher odds of graft dysfunction.

