---
title: "Donor Glucose and Delayed Graft Dysfunction in Liver Transplantation"
subtitle: "Does Donor Risk Index Modify the Association?"
date: "STA 198, Spring 2025"
names: Kuber Gohil, Elena Maeve, Luke Wang

output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)

DonorHyperglycemia <- read_csv("DonorHyperglycemia.csv")

m1 <- glm(delayed_fn ~ log2(glutwa) * donorrisk + d_genderf,
          data = DonorHyperglycemia, family = "binomial")

m0 <- glm(delayed_fn ~ log2(glutwa) + donorrisk + d_genderf,
          data = DonorHyperglycemia, family = "binomial")

m <- glm(delayed_fn ~ donorrisk + d_genderf,
         data = DonorHyperglycemia, family = "binomial")
```

# Introduction

Delayed graft dysfunction (DGD) is a complication that can occur following liver
transplantation, and is generally associated with longer hospital stays, higher
mortality, and worse long-term graft survival. Understanding what predicts DGD
matters both for patient outcomes and for decisions about which donor organs to
accept.

One commonly used tool for assessing donor organ quality is the Donor Risk Index
(DRI), a composite score that takes into account factors like donor age, cause of
death, and organ type to estimate the likelihood of graft failure. Higher DRI
values indicate higher-risk donations. While DRI has been used to predict
transplant outcomes, it is based on demographic and clinical information and does
not include real-time physiological measurements from the donor.

Donor glucose is one variable that might provide additional information. High
glucose levels in critically ill patients can be a sign of physiological stress,
which could potentially affect organ viability before transplantation. That said,
whether donor glucose is actually associated with DGD — and whether that
relationship changes depending on how risky the donor already is by DRI — is not
well established in the literature we reviewed.

This study looks at both of these questions using data from 572 liver transplant
donors. The dataset includes donor glucose levels (recorded as a time-weighted
average and log-transformed for analysis), the Donor Risk Index, donor sex, and
whether delayed graft dysfunction occurred.

**Research question:** Does donor glucose level predict delayed graft dysfunction
in liver transplantation, and does this association differ by donor risk index?

# Methodology

## Model Specification

Because delayed graft dysfunction is a binary outcome (1 = dysfunction,
0 = no dysfunction), we used logistic regression to model the probability of DGD.
This is a standard approach for binary outcomes and lets us estimate how each
predictor is associated with the odds of dysfunction while holding the others
constant.

We fit two models. The first includes an interaction term between log-transformed
glucose and DRI. We included this interaction because we wanted to test whether
the relationship between glucose and DGD might depend on how risky the donor
already is — for example, glucose might matter more for high-risk donors than
low-risk ones:

$$\text{logit}(P(\text{DGD})) = \beta_0 + \beta_1\log_2(\text{glucose}) + \beta_2\text{DRI} + \beta_3\text{sex} + \beta_4\log_2(\text{glucose})\times\text{DRI}$$

where $\text{logit}(P(\text{DGD}))$ is the log-odds of delayed graft dysfunction,
$\log_2(\text{glucose})$ is the log base-2 transformed time-weighted average donor
glucose, $\text{DRI}$ is the Donor Risk Index, $\text{sex}$ is donor sex
(female = 1), and the interaction term $\log_2(\text{glucose}) \times \text{DRI}$
captures whether the glucose effect differs across DRI levels.

We also fit a reduced model without the interaction term, to look at the
independent effect of each predictor:

$$\text{logit}(P(\text{DGD})) = \beta_0 + \beta_1\log_2(\text{glucose}) + \beta_2\text{DRI} + \beta_3\text{sex}$$

We then compared the two models using a likelihood ratio test (LRT) to see
whether adding the interaction term meaningfully improved model fit. If the test
came back non-significant ($p > 0.05$), we would conclude there is not enough
evidence for an interaction and stick with the simpler model.

## Choice of Predictors and Transformations

Glucose was log-transformed (base 2) before modeling because the raw values were
right-skewed. The log$_2$ scale is also convenient for interpretation: a one-unit
increase corresponds to a doubling of the original glucose value.

We included donor sex as a covariate based on prior literature suggesting it may
be related to graft outcomes. DRI was kept as a continuous variable rather than
grouped into categories, since collapsing it would lose information.

## Assumptions

Logistic regression requires a few key assumptions: that the log-odds of the
outcome is approximately linear in each continuous predictor, that observations
are independent, and that predictors are not perfectly collinear with each other.

We assessed linearity by looking at the shape of the fitted logistic curve, which
appeared smooth and monotonic — consistent with the linearity assumption, though
we did not run formal diagnostics. Independence seems reasonable since each row
in the dataset is a separate donor-recipient pair. The three predictors capture
different aspects of the donor (metabolic, composite risk, and demographic), so
perfect collinearity is unlikely. It is worth noting that 135 observations were
dropped due to missing data, leaving 437 complete cases, which is still a
reasonable sample size for this kind of model.

# Results

## Model Comparison: Likelihood Ratio Test

The likelihood ratio test comparing the full model to the reduced model returned
a $p$-value of 0.32. This means we do not have sufficient evidence at the 0.05
significance level to conclude that the interaction between glucose and DRI
improves model fit. In other words, the data do not support the idea that the
effect of glucose on DGD changes depending on the donor's risk level. We
proceeded with the reduced model for the rest of the analysis.

```{r lrt-table}
lrt_df <- data.frame(
  Model     = c("Reduced (no interaction)", "Full (with interaction)"),
  Resid_df  = c(433, 432),
  Resid_Dev = c(492.50, 491.51),
  Delta_Dev = c("—", "0.988"),
  p_value   = c("—", "0.320")
)

kable(lrt_df,
      booktabs  = TRUE,
      col.names = c("Model", "Resid. df", "Resid. Dev.", "$\\Delta$ Dev.", "$p$-value"),
      escape    = FALSE,
      caption   = "Likelihood Ratio Test: Full vs. Reduced Model") |>
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) |>
  footnote(general = "Null deviance = 503.72 on 436 df. 135 observations removed due to missingness.",
           general_title = "", footnote_as_chunk = TRUE)
```

## Reduced Model Coefficients

Looking at the reduced model, donor risk index was the only statistically
significant predictor (estimate = 0.958, SE = 0.328, $z$ = 2.92, $p$ = 0.003).
Donor glucose ($p$ = 0.282) and donor sex ($p$ = 0.878) were not significant at
the 0.05 level.

```{r coeff-table}
tidy(m0) |>
  mutate(
    term = c("(Intercept)", "$\\log_2$(glucose)", "donorrisk", "sex (female)"),
    p.value = case_when(
      p.value < 0.01 ~ paste0(round(p.value, 3), " **"),
      p.value < 0.05 ~ paste0(round(p.value, 3), " *"),
      TRUE           ~ as.character(round(p.value, 3))
    ),
    across(where(is.numeric), ~round(., 3))
  ) |>
  kable(booktabs  = TRUE,
        escape    = FALSE,
        col.names = c("Variable", "Estimate", "Std. Error", "$z$ value", "$p$-value"),
        caption   = "Reduced Logistic Regression Model: Coefficients") |>
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) |>
  row_spec(3, bold = TRUE) |>
  footnote(general = "** $p < 0.01$, * $p < 0.05$. AIC = 500.5.",
           general_title = "", escape = FALSE, footnote_as_chunk = TRUE)
```

## Effect Size and Confidence Interval for Donor Risk

To better understand the size of the DRI effect, we exponentiated the coefficient
to get an odds ratio of approximately 2.61 (95% CI: 1.40 to 5.07). This suggests
that for each one-unit increase in DRI, the estimated odds of delayed graft
dysfunction increase by about 161%, holding the other variables constant. The
confidence interval is fairly wide, which reflects some uncertainty in the
estimate, but it lies entirely above 1.0 — suggesting the positive association
is reasonably consistent.

```{r or-table}
cbind(OR = exp(coef(m0)), exp(confint(m0))) |>
  as.data.frame() |>
  round(3) |>
  tibble::rownames_to_column("Variable") |>
  mutate(Variable = c("(Intercept)", "$\\log_2$(glucose)", "donorrisk", "sex (female)")) |>
  kable(booktabs  = TRUE,
        escape    = FALSE,
        col.names = c("Variable", "Odds Ratio", "2.5\\% CI", "97.5\\% CI"),
        caption   = "Odds Ratios and 95\\% Confidence Intervals (Reduced Model)") |>
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) |>
  row_spec(3, bold = TRUE) |>
  footnote(general = "Odds ratios = exp(coef(m0)). CIs via profiled log-likelihood.",
           general_title = "", footnote_as_chunk = TRUE)
```

## Visualization

Figure 1 shows the predicted probability of delayed graft dysfunction across the
range of observed DRI values. The upward trend is fairly clear — donors with
higher risk scores are predicted to have a higher probability of DGD. The
confidence band is tighter at lower DRI values where there are more observations,
and widens toward the upper end where data are sparser.

```{r fig1, fig.cap="Predicted probability of delayed graft dysfunction as a function of Donor Risk Index, with 95\\% confidence band. Fitted from a binomial GLM using geom\\_smooth().", fig.align="center", fig.width=5.5, fig.height=3.8, out.width="75%"}

# ── PASTE YOUR GGPLOT CODE HERE ──────────────────────────────────────────────
ggplot(DonorHyperglycemia, aes(x = donorrisk, y = delayed_fn)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(x = "Donor Risk Index", y = "Probability of Graft Dysfunction") +
  theme_classic()
# ─────────────────────────────────────────────────────────────────────────────
```

# Discussion

Overall, our results suggest that donor risk index is a meaningful predictor of
delayed graft dysfunction, while donor glucose does not appear to add much
on top of what DRI already captures — at least in this dataset and with this
modeling approach.

The odds ratio for DRI was about 2.61, which seems like a substantial effect,
though we want to be careful about over-interpreting a single model with limited
covariates. The confidence interval (1.40 to 5.07) is wide enough that there is
real uncertainty around this estimate. Still, the direction is clear and the
result is statistically significant, which gives us some confidence that DRI is
picking up on something real.

The non-significance of glucose was somewhat surprising given our initial
hypothesis, but there are a few reasonable explanations. It is possible that DRI
already captures much of what glucose reflects — donors with high DRI scores tend
to be older and sicker, which could also drive elevated glucose. If that is the
case, glucose may not tell us much beyond what DRI already does. It is also
possible that a glucose effect exists but that our sample of 437 complete cases
was not large enough to detect it reliably, especially after 135 observations were
dropped due to missingness. We cannot really distinguish between these two
explanations with the data we have.

There are a few other limitations worth mentioning. Our model only includes three
predictors, and there are likely other variables — such as cold ischemia time or
recipient-level factors — that could matter and are not accounted for here. Because
this is observational data, we also cannot make any causal claims. A higher DRI
being associated with more dysfunction does not mean DRI causes it; both could be
driven by underlying donor characteristics we are not measuring. We also handled
missing data by dropping incomplete cases, which is simple but may introduce bias
if the missingness is related to the outcome.

If we were to extend this analysis, multiple imputation could help recover some of
the information lost from the 135 dropped observations. Including additional
clinical variables would also likely improve the model. It would also be
interesting to look at glucose at different time points rather than just a
time-weighted average, to see if timing matters.

# Conclusion

We used logistic regression to examine whether donor glucose predicts delayed graft
dysfunction and whether that relationship is modified by donor risk index. The
interaction between glucose and DRI was not statistically significant (LRT
$p$ = 0.32), and glucose itself was not a significant predictor in the reduced
model ($p$ = 0.28). Donor risk index, on the other hand, showed a consistent
positive association with DGD across both models, with an estimated odds ratio of
2.61 (95% CI: 1.40 to 5.07).

Based on these results, DRI appears to be the stronger predictor of graft
dysfunction in this dataset, and glucose did not provide meaningful additional
information beyond it. That said, our model is relatively simple and there is
quite a bit of uncertainty in some of our estimates, so these findings should be
interpreted cautiously rather than as definitive conclusions.
We are 95% confident that a one-unit increase in donor risk is associated with between 1.4× and 5.1× higher odds of graft dysfunction.

